# (APPENDIX) Appendix {-}

<!-- Please don't mess with the next few lines! -->
<style>h5{font-size:2em;color:#0000FF}h6{font-size:1.5em;color:#0000FF}div.answer{margin-left:5%;border:1px solid #0000FF;border-left-width:10px;padding:25px} div.summary{background-color:rgba(30,144,255,0.1);border:3px double #0000FF;padding:25px}</style>`r options(scipen=999)`<p style="color:#ffffff">`r intToUtf8(c(50,46,48))`</p>
<!-- Please don't mess with the previous few lines! -->

# Rubric for inference {#appendix-rubric}

This is the R Markdown outline for running inference, both a hypothesis test and a confidence interval.


## Exploratory data analysis {-}

### Use data documentation (help files, code books, Google, etc.) to determine as much as possible about the data provenance and structure. {-}


::: {.answer}

Please write up your answer here

```{r}
# Add code here to print the data
```

```{r}
# Add code here to glimpse the variables
```

:::

### Prepare the data for analysis. [Not always necessary.] {-}

::: {.answer}

```{r}
# Add code here to prepare the data for analysis.
```

:::

### Make tables or plots to explore the data visually. {-}

::: {.answer}

```{r}
# Add code here to make tables or plots.
```

:::


## Hypotheses {-}

### Identify the sample (or samples) and a reasonable population (or populations) of interest. {-}

::: {.answer}

Please write up your answer here.

:::

### Express the null and alternative hypotheses as contextually meaningful full sentences. {-}

::: {.answer}

$H_{0}:$ Null hypothesis goes here.

$H_{A}:$ Alternative hypothesis goes here.

:::

### Express the null and alternative hypotheses in symbols (when possible). {-}

::: {.answer}

$H_{0}: math$

$H_{A}: math$

:::


## Model {-}

### Identify the sampling distribution model. {-}

::: {.answer}

Please write up your answer here.

:::

### Check the relevant conditions to ensure that model assumptions are met. {-}

::: {.answer}

Please write up your answer here. (Some conditions may require R code as well.)

:::


## Mechanics {-}

### Compute the test statistic. {-}

::: {.answer}

```{r}
# Add code here to compute the test statistic.
```

:::

### Report the test statistic in context (when possible). {-}

::: {.answer}

Please write up your answer here.

:::

### Plot the null distribution. {-}

::: {.answer}

```{r}
set.seed(1)
# Add code here to simulate the null distribution.
# Run 1000 reps like in the earlier example.
```

```{r}
# Add code here to plot the null distribution.
```

:::

### Calculate the P-value. {-}

::: {.answer}

```{r}
# Add code here to calculate the P-value.
```

:::

### Interpret the P-value as a probability given the null. {-}

::: {.answer}

Please write up your answer here.

:::


## Conclusion {-}

### State the statistical conclusion. {-}

::: {.answer}

Please write up your answer here.

:::

### State (but do not overstate) a contextually meaningful conclusion. {-}

::: {.answer}

Please write up your answer here.

:::

### Express reservations or uncertainty about the generalizability of the conclusion. {-}

::: {.answer}

Please write up your answer here.

:::

### Identify the possibility of either a Type I or Type II error and state what making such an error means in the context of the hypotheses. {-}

::: {.answer}

Please write up your answer here.

:::

## Confidence interval {-}

### Check the relevant conditions to ensure that model assumptions are met.  {-}

::: {.answer}

Please write up your answer here. (Some conditions may require R code as well.)

:::

### Calculate the confidence interval. {-}

::: {.answer}

```{r}
# Add code here to calculate the confidence interval.
```

:::

### State (but do not overstate) a contextually meaningful interpretation. {-}

::: {.answer}

Please write up your answer here.

:::

### If running a two-sided test, explain how the confidence interval reinforces the conclusion of the hypothesis test. {-}

::: {.answer}

Please write up your answer here.

:::

### If the P-value is statistically significant and the confidence interval does not contain zero, comment on the effect size and the practical significance of the result. [Not always relevant.] {-}

::: {.answer}

Please write up your answer here.

:::

