# (APPENDIX) Appendix {-}

# Rubric for inference {#appendix-rubric}

This is the R Markdown outline for running inference, both a hypothesis test and a confidence interval.


## Exploratory data analysis  {#appendix-rubric-eda}

### Use data documentation (help files, code books, Google, etc.) to determine as much as possible about the data provenance and structure. {#appendix-rubric--documentation}


::: {.answer}

Please write up your answer here

```{r}
# Add code here to print the data
```

```{r}
# Add code here to glimpse the variables
```

:::

### Prepare the data for analysis. [Not always necessary.] {#appendix-rubric-prepare}

::: {.answer}

```{r}
# Add code here to prepare the data for analysis.
```

:::

### Make tables or plots to explore the data visually. {#appendix-rubric-plots}

::: {.answer}

```{r}
# Add code here to make tables or plots.
```

:::


## Hypotheses {#appendix-rubric-hypotheses}

### Identify the sample (or samples) and a reasonable population (or populations) of interest. {#appendix-rubric-sample-pop}

::: {.answer}

Please write up your answer here.

:::

### Express the null and alternative hypotheses as contextually meaningful full sentences. {#appendix-rubric-express-words}

::: {.answer}

$H_{0}:$ Null hypothesis goes here.

$H_{A}:$ Alternative hypothesis goes here.

:::

### Express the null and alternative hypotheses in symbols (when possible). {#appendix-rubric-express-math}

::: {.answer}

$H_{0}: math$

$H_{A}: math$

:::


## Model {#appendix-rubric-model}

### Identify the sampling distribution model. {#appendix-rubric-sampling-dist-model}

::: {.answer}

Please write up your answer here.

:::

### Check the relevant conditions to ensure that model assumptions are met. {#appendix-rubric-ht-conditions}

::: {.answer}

Please write up your answer here. (Some conditions may require R code as well.)

:::


## Mechanics {#appendix-rubric-mechanics}

### Compute the test statistic. {#appendix-rubric-compute-test-stat}

::: {.answer}

```{r}
# Add code here to compute the test statistic.
```

:::

### Report the test statistic in context (when possible). {#appendix-rubric-report-test-stat}

::: {.answer}

Please write up your answer here.

:::

### Plot the null distribution. {#appendix-rubric-plot-null}

::: {.answer}

```{r}
set.seed(1)
# Add code here to simulate the null distribution.
# Run 1000 reps like in the earlier example.
```

```{r}
# Add code here to plot the null distribution.
```

:::

### Calculate the P-value. {#appendix-rubric-calculate-p}

::: {.answer}

```{r}
# Add code here to calculate the P-value.
```

:::

### Interpret the P-value as a probability given the null. {#appendix-rubric-interpret-p}

::: {.answer}

Please write up your answer here.

:::


## Conclusion {#appendix-rubric-ht-conclusion}

### State the statistical conclusion. {#appendix-rubric-stat-conclusion}

::: {.answer}

Please write up your answer here.

:::

### State (but do not overstate) a contextually meaningful conclusion. {#appendix-rubric-context-conclusion}

::: {.answer}

Please write up your answer here.

:::

### Express reservations or uncertainty about the generalizability of the conclusion. {#appendix-rubric-reservations}

::: {.answer}

Please write up your answer here.

:::

### Identify the possibility of either a Type I or Type II error and state what making such an error means in the context of the hypotheses. {#appendix-rubric-errors}

::: {.answer}

Please write up your answer here.

:::

## Confidence interval {#appendix-rubric-ci}

### Check the relevant conditions to ensure that model assumptions are met. {#appendix-rubric-ci-conditions}

::: {.answer}

Please write up your answer here. (Some conditions may require R code as well.)

:::

### Calculate the confidence interval. {#appendix-rubric-calculate-ci}

::: {.answer}

```{r}
# Add code here to calculate the confidence interval.
```

:::

### State (but do not overstate) a contextually meaningful interpretation. {#appendix-rubric-ci-conclusion}

::: {.answer}

Please write up your answer here.

:::

### If running a two-sided test, explain how the confidence interval reinforces the conclusion of the hypothesis test. {#appendix-rubric-ci-reinforces-ht}

::: {.answer}

Please write up your answer here.

:::

### If the P-value is statistically significant and the confidence interval does not contain zero, comment on the effect size and the practical significance of the result. [Not always relevant.] {#appendix-rubric-effect-size}

::: {.answer}

Please write up your answer here.

:::

