---
title: "Rubric for Inference"
output: pdf_document
---
<!-- Please don't mess with the next few lines! -->
\def\hrulefill{\leavevmode\leaders\hrule height 2pt\hfill\kern 0pt}
\newcommand{\answerbegin}{\par\textcolor{blue}{\hrulefill\quad ANSWER\quad\hrulefill\newline}}
\newcommand{\answerend}{\par\textcolor{blue}{\hrulefill}\newline}
`r options(scipen = 999)`
<!-- Please don't mess with the previous few lines! -->


This is the R Markdown outline for running inference, both a hypothesis test and a confidence interval.


## Exploratory data analysis

### Use data documentation (help files, code books, Google, etc.), the str command, and other summary functions to understand the data.

\answerbegin
```{r}
# Add code here to understand the data.
```
\answerend

### Prepare the data for analysis. [Not always necessary.]

\answerbegin
```{r}
# Add code here to prepare the data for analysis.
```
\answerend

### Make tables or plots to explore the data visually.

\answerbegin
```{r}
# Add code here to make tables or plots.
```
\answerend


## Hypotheses

### Identify the sample (or samples) and a reasonable population (or populations) of interest.

\answerbegin
Please write up your answer here.
\answerend

### Express the null and alternative hypotheses as contextually meaningful full sentences.

\answerbegin
$H_{0}$: Null hypothesis goes here.

$H_{A}$: Alternative hypothesis goes here.
\answerend

### Express the null and alternative hypotheses in symbols (when possible).

\answerbegin
$H_{0}: math$

$H_{A}: math$
\answerend


## Model

### Identify the sampling distribution model.

\answerbegin
Please write up your answer here.
\answerend

### Check the relevant conditions to ensure that model assumptions are met.

\answerbegin
Please write up your answer here. (Some conditions may require R code as well.)
\answerend


## Mechanics

### Compute the test statistic.

\answerbegin
```{r}
# Add code here to compute the test statistic.
```
\answerend

### Plot the null distribution.

\answerbegin
```{r}
# Add code here to plot the null distribution.
```
\answerend

### Calculate the P-value.

\answerbegin
```{r}
# Add code here to calculate the P-value.
```
\answerend


## Conclusion

### State the statistical conclusion.

\answerbegin
Please write up your answer here.
\answerend

### State (but do not overstate) a contextually meaningful conclusion.

\answerbegin
Please write up your answer here.
\answerend

### Identify the possibility of either a Type I or Type II error and state what making such an error means in the context of the hypotheses.

\answerbegin
Please write up your answer here.
\answerend


## Confidence interval

### Check the relevant conditions to ensure that model assumptions are met.

\answerbegin
Please write up your answer here. (Some conditions may require R code as well.)
\answerend

### Calculate the confidence interval.

\answerbegin
```{r}
# Add code here to calculate the confidence interval.
```
\answerend

### State (but do not overstate) a contextually meaningful interpretation.

\answerbegin
Please write up your answer here.
\answerend
