---
title: "Rubric for Inference"
output: pdf_document
---
<!-- Please don't mess with the next two lines! -->
\newenvironment{answer}{\definecolor{shadecolor}{RGB}{225, 225, 255}\begin{shaded}}{\end{shaded}}
<!-- Please don't mess with the previous two lines! -->


This is the R Markdown outline for running inference. For convenience, both the rubric for inference with simulation and inference with a sampling distribution model are included. Following these is the rubric for confidence intervals.

# Rubric for inference with simulation:


## Exploratory data analysis

### Use data documentaton (help files, code books, Google, etc.), the str command, and other summary functions to understand the data.

```{r}
## Add code here to understand the data.
```

### Prepare the data for analysis. [Not always necessary.]

```{r}
## Add code here to prepare the data for analysis.
```

### Make tables or plots to explore data visually.

```{r}
## Add code here to make tables or plots.
```


## Hypotheses

### Identify the sample (or samples) and a reasonable population (or populations) of interest.

\begin{answer}
Please write up your answer here.
\end{answer}

### Express the null and alternative hypotheses as contextually meaningful full sentences.

\begin{answer}
$H_0:$ Null hypothesis goes here.

$H_A:$ Alternative hypothesis goes here.
\end{answer}

### Express the null and alternative hypotheses in symbols.

\begin{answer}
$H_0: math$

$H_A: math$
\end{answer}


## Model

### Check the relevant conditions to ensure that the assumptions are met.

\begin{answer}
Please write up your answer here.
\end{answer}


## Mechanics

### Compute the test statistic.

```{r}
## Add code here to compute the test statistic.
```

### Plot simulated values of the null distribution.

```{r}
## Add code here to plot simulated values of the null distribution.
```

### Calculate the P-value.

```{r}
## Add code here to calculate the P-value.
```


## Conclusion

### State the statistical conclusion.

\begin{answer}
Please write up your answer here.
\end{answer}

### State (but do not overstate) a contextually meaningful conclusion.

\begin{answer}
Please write up your answer here.
\end{answer}

### Identify the possibility of either a Type I or Type II error and state what making such an error means in the context of the hypotheses.

\begin{answer}
Please write up your answer here.
\end{answer}


\newpage


# Rubric for inference with a sampling distribution model:


## Exploratory data analysis

### Use data documentaton (help files, code books, Google, etc.), the str command, and other summary functions to understand the data.

```{r}
## Add code here to understand the data.
```

### Prepare the data for analysis. [Not always necessary.]

```{r}
## Add code here to prepare the data for analysis.
```

### Make tables or plots to explore data visually.

```{r}
## Add code here to make tables or plots.
```


## Hypotheses

### Identify the sample (or samples) and a reasonable population (or populations) of interest.

\begin{answer}
Please write up your answer here.
\end{answer}

### Express the null and alternative hypotheses as contextually meaningful full sentences.

\begin{answer}
$H_0$: Null hypothesis goes here.

$H_A$: Alternative hypothesis goes here.
\end{answer}

### Express the null and alternative hypotheses in symbols.

\begin{answer}
$H_0: math$

$H_A: math$
\end{answer}


## Model

### Check the relevant conditions to ensure that the assumptions are met.

\begin{answer}
Please write up your answer here.
\end{answer}

(Some conditions may also require R code chunks to check.)


## Mechanics

### Compute the test statistic.

```{r}
## Add code here to compute the test statistic.
```

### Plot the null distribution.

```{r}
## Add code here to plot the null distribution.
```

### Calculate the P-value.

```{r}
## Add code here to calculate the P-value.
```


## Conclusion

### State the statistical conclusion.

\begin{answer}
Please write up your answer here.
\end{answer}

### State (but do not overstate) a contextually meaningful conclusion.

\begin{answer}
Please write up your answer here.
\end{answer}

### Identify the possibility of either a Type I or Type II error and state what making such an error means in the context of the hypotheses.

\begin{answer}
Please write up your answer here.
\end{answer}


\newpage


## Confidence interval

### Conditions

\begin{answer}
Please write up your answer here.
\end{answer}

### Calculation

```{r}
## Add code here to calculate the confidence interval.
```

### Conclusion

\begin{answer}
Please write up your answer here.
\end{answer}
