---
title: "Rubric for Inference"
author: "Sean Raleigh"
date: "October 31, 2015"
output: html_document
---

This is the R Markdown outline for running inference. For convenience, both the rubric for inference with simulation and inference with a sampling distribution model are included. Following these is the rubric for confidence intervals.

# Rubric for inference with simulation:

## Hypotheses

### Identify the sample and a reasonable population of interest.

<!-- Please write up your answer in the space below -->


### Express the null and alternative hypotheses as contextually meaningful full sentences.

<!-- Please write up your answer in the space below -->


### Express the null and alternative hypotheses in symbols.

<!-- Please write up your answer in the space below -->


## Model

### Identify the correct sampling distribution model.

This step is not applicable.

### Check the relevant conditions to ensure that the assumptions are met.

<!-- Please write up your answer in the space below -->


## Mechanics

### Compute the test statistic.

```{r}
## Add code here to compute the test statistic.
```

### Plot simulated values of the null distribution.

```{r}
## Add code here to plot simulated values of the null distribution.
```

### Calculate the P-value.

```{r}
## Add code here to calculate the P-value.
```


## Conclusion

### State the statistical conclusion.

<!-- Please write up your answer in the space below -->


### State (but do not overstate) a contextually meaningful conclusion.

<!-- Please write up your answer in the space below -->


### Identify the possibility of either a Type I or Type II error and state what making such an error means in the context of the hypotheses.

<!-- Please write up your answer in the space below -->


# Rubric for inference with a sampling distribution model:


## Hypotheses

### Identify the sample and a reasonable population of interest.

<!-- Please write up your answer in the space below -->


### Express the null and alternative hypotheses as contextually meaningful full sentences.

<!-- Please write up your answer in the space below -->


### Express the null and alternative hypotheses in symbols.

<!-- Please write up your answer in the space below -->


## Model

### Identify the correct sampling distribution model.

<!-- Please write up your answer in the space below -->


### Check the relevant conditions to ensure that the model assumptions are met.

<!-- Please write up your answer in the space below -->


## Mechanics

### Compute the test statistic.

```{r}
## Add code here to compute the test statistic.
```

### Plot the null distribution.


```{r}
## Add code here to plot the null distribution.
```

### Calculate the P-value.

```{r}
## Add code here to calculate the P-value.
```


## Conclusion

### State the statistical conclusion.

<!-- Please write up your answer in the space below -->


### State (but do not overstate) a contextually meaningful conclusion.

<!-- Please write up your answer in the space below -->


### Identify the possibility of either a Type I or Type II error and state what making such an error means in the context of the hypotheses.

<!-- Please write up your answer in the space below -->


## Confidence interval

### Conditions

<!-- Please write up your answer in the space below -->


### Calculation

```{r}
## Add code here to calculate the confidence interval.
```

### Conclusion

<!-- Please write up your answer in the space below -->
