---
title: "Summary statistics"
author: "Put your name here"
date: "Put the date here"
output: pdf_document
---
<!-- Please don't mess with the next few lines! -->
\def\hrulefill{\leavevmode\leaders\hrule height 2pt\hfill\kern 0pt}
\newcommand{\answerbegin}{\par\textcolor{blue}{\hrulefill\quad ANSWER\quad\hrulefill\newline}}
\newcommand{\answerend}{\par\textcolor{blue}{\hrulefill}\newline}
`r options(scipen = 999)`
<!-- Please don't mess with the previous few lines! -->


## Introduction

In this module, we'll learn about summary statistics, which are numerical summaries calculated from data.


## Instructions

Presumably, you have already created a new project and downloaded this file into it. From the `Run` menu above, select `Run All` to run all existing code chunks.

Sometimes you will be asked to add your own R code. That will appear in this document as a code chunk with a request for you to add your own code, like so:

```{r}
# Add code here
```

Be sure to remove the line `# Add code here` when you have added your own code. You should run each new code chunk you create by clicking on the dark green arrow in the upper-right corner of the code chunk.

Sometimes you will be asked to type up your thoughts. That will appear in the document as follows:

\answerbegin
Please write up your answer here.
\answerend

Be sure to remove the line "Please write up your answer here" when you have written up your answer. In these areas of the assignment, please use contextually meaningful full sentences/paragraphs (unless otherwise indicated) and proper spelling, grammar, punctuation, etc. This is not R code, but rather a free response section where you talk about your analysis and conclusions. If you need to use some R code as well, you can use inline R code or start a new code chunk.

When you are finished with the assignment, knit to PDF and proofread the PDF file **carefully**. Do not download the PDF file from the PDF viewer; rather, you should export the PDF file to your computer by selecting the check box next to the PDF file in the Files plane, clicking the `More` menu, and then clicking `Export`. Submit your assignment according to your professor's instructions.


## Load Packages

We load the `mosaic` package as well as the `MASS` package so that we can work with data on risk factors associated with low birth weight.

```{r, warning = FALSE, message=FALSE}
library(mosaic)
library(MASS)
```


## A note about mathematical notation

From time to time, we will use mathematical notation that can't be typed directly on the keyboard. For example, let's suppose we want to typeset the quadratic formula, which involves a complicated fraction as well as a square root symbol.

When such notation appears, it will be surrounded by double dollar signs as follows:

$$ x = \frac{-b \pm \sqrt{b^{2} -  4ac}}{2a} $$

The Markdown document will interpret this special mathematical notation and render it on the screen as well as the PDF document.^[This notation is part of a mathematical document preparation system called LaTeX, pronounced "Lay-tek" (not like the rubbery substance).]  If the nicely formatted formula does not appear on your screen, place your cursor anywhere inside the math formula and hit Ctrl-Enter or Cmd-Enter (PC or Mac respectively).

Sometimes, we want such math to appear inline. We can do this with single dollar signs: $(x_{2} - x_{1})^{2} + (y_{2} -  y_{1})^{2} = r^{2}$.

This will *not* render visually in the Markdown document, but it will show up in the PDF file. If you want to check that it worked properly without having to Knit to PDF, you can either hover your cursor over the math formula and wait a second, or you can place your cursor anywhere inside the math formula and hit Ctrl-Enter or Cmd-Enter (PC or Mac respectively) to see a pop-up window previewing the mathematical content properly formatted.

You will be shown examples of any mathematical notation you need to use in any given assignment, so feel free to copy/paste/modify any math notation you need.


## Mean and standard deviation

Summary statistics come in two general flavors: measures of center and measures of spread.

The first pair we'll consider is the mean and the standard deviation. The *mean*---denoted $\bar{y}$---of a variable $y$ is calculated by summing all the values of the variable, and dividing by the total number of observations. In formula form, this is

$$\bar{y} = \frac{\sum y}{n}.$$

This is a measure of center since it estimates the "middle" of a set of numbers.

It is calculated in R using the `mean` command. For example, if we want to calculate the mean weight in pounds of the mother at the last menstrual period, we type the following:

```{r}
mean(birthwt$lwt)
```

The corresponding measure of spread is the *standard deviation*. Usually this is called $s$ and is calculated using a much more complicated formula:

$$s = \sqrt{\frac{\sum (y - \bar{y})^2}{n - 1}}.$$

This is a measure of spread because the $(y - \bar{y})$ term measures the how far away each data point is from the mean.

In R, this is calculated with the `sd` command.

```{r}
sd(birthwt$lwt)
```

The mean and the standard deviation should always be reported together.

Another related measurement is the *variance*, but this is nothing more than the standard deviation squared:

$$s^2 = \frac{\sum (y - \bar{y})^2}{n - 1}.$$

(Compare this formula to the one for the standard deviation. Nothing has changed except for the removal of the square root.) We rarely use the variance in an introductory stats class because it's not as interpretable as the standard deviation.^[The main reason for this is units. If the data variable is the mother's weight in pounds, then both the mean and the standard deviation are also reported in pounds. The variance has units of "pounds squared", but what does that even mean?] If you need to do this in R, the command is `var`.

```{r}
var(birthwt$lwt)
```

(You can check and see that the number above really is just `r sd(birthwt$lwt)` squared. Remember, in the R Markdown document, you can click inside the inline R code chunk from the previous sentence and hit Ctrl-Enter or Cmd-Enter.)


## Median and IQR

Another choice for measuring the center and spread of a data set is the median and the IQR. The median is just the middle value if the list of values is ordered. In R, it is calculated using the `median` command.

```{r}
median(birthwt$lwt)
```

You can check that this is correct: if we print out the entire `lwt` variable, you can see all 189 values, and if we're clever about it, we can see them in order.

```{r}
sort(birthwt$lwt)
```

### Exercise

If there are 189 mothers in this data set, in which position in the list will the median value appear? (Be careful: you can't just divide 189 by 2,  although if you do, you'll be close to the right answer.) Verify that the median value 121 does appear in the position you calculated.

\answerbegin
Please write up your answer here.
\answerend

* * * * *

Calculating the *interquartile range*---or *IQR*---requires first the calculation of the first and third quartiles, denoted Q1 and Q3. If the median is the 50% mark in the sorted data, the first and third quartiles are the 25% and the 75% marks, respectively.^[One way to compute these is to calculate the median of the lower and upper halves of the data separately. Then again, it's hard to know how to split the data set into halves if there are an odd number of observations. There are many different methods for computing percentiles in general, but you don't need to worry too much about the particular implementation in R.] Then the IQR is just

$$IQR = Q3 - Q1$$

In R, you can get the IQR by using---are you ready for this?---the `IQR` command.

```{r}
IQR(birthwt$lwt)
```

The IQR is a measure of spread because the distance between Q1 and Q3 measures the span of the "middle 50%" of the data.

### Exercise

A general function for computing any percentile in R is the `quantile` function. For example, since Q1 is the 25th percentile, you can compute it as follows:

```{r}
Q1 <- quantile(birthwt$lwt, 0.25)
unname(Q1)
```

(The `unname` command just helps clean up the output here. Don't worry too much about it.)

Now compute Q3. Also check that the IQR calculated above matches the value you get from subtracting Q3 minus Q1.

\answerbegin
```{r}
# Add code here to compute and print out Q3.
# You should use the unname command as in the above chunk.
```

```{r}
# Add code here to compute Q3 - Q1
```
\answerend

*****

The median and the IQR should always be reported together.


## Robust statistics

Some statistics are more sensitive than others to features of the data. For example, outliers are data points that are far away from the bulk of the data. The mean and especially the standard deviation can change a lot when outliers are present. Also, skewness in the data frequently pulls the mean too far in the direction of the skew while simultaneously inflating the standard deviation.

On the other hand, the median and IQR are "robust", meaning that they do not change much (or at all) in the presence of outliers and they tend to be good summaries even for skewed data.

### Exercise

Explain why the median and IQR are robust. In other words, why does an outlier have little or no influence on the median and IQR?

\answerbegin
Please write up your answer here.
\answerend

*****


## Five-number summary

A *five-number summary* is the minimum, Q1, median, Q3, and maximum of a set of numbers.

The `summary` command in R gives you the five-number summary, and throws in the mean for good measure.

```{r}
summary(birthwt$lwt)
```

You can, of course, isolate the various pieces of this. You already know most of the commands below.

```{r}
min(birthwt$lwt)
```
```{r}
median(birthwt$lwt)
```
```{r}
max(birthwt$lwt)
```
```{r}
mean(birthwt$lwt)
```
```{r}
quantile(birthwt$lwt)
```

### Exercise

What is the difference between the way `quantile` was used in a previous exercise versus the way it was used here? How did that change the output?

\answerbegin
Please write up your answer here.
\answerend

*****

The `mosaic` package also has a summary command called `favstats` that has a little more information, including the standard deviation, the sample size, and a count of any cases that are missing data. This command is much more useful than the standard `summary` command.

```{r}
favstats(birthwt$lwt)
```


Also, don't forget about the trick for using R commands inline. If you need to mention a statistic in the middle of a sentence, there is no need to break the sentence and display a code chunk. Be sure you're looking at the R Markdown document to note that the numbers in the next sentence are not manually entered, but are calculated on the fly:

There are `r NROW(birthwt)` births represented in this data and the median weight of the women as of their last menstrual period is `r median(birthwt$lwt)` pounds.

### Your turn

Type a full, contextually meaningful sentence using inline R code (as above) summarizing the minimum and maximum baby weights (in grams) in our data set.

\answerbegin
Please write up your answer here.
\answerend


## Summary statistics by group

Using base R, it's not so easy to get summary statistics for each group separately. Fortunately, the `mosaic` package comes to the rescue, allowing for more flexibility.

For example:

```{r}
favstats(lwt ~ race, data = birthwt)
```

The notation is a little weird. Don't worry about the tilde for now. Just learn that it needs to be there. This says, "Look at the numerical variable `lwt` grouped by `race`." Indeed, the output has three lines, one for each race in the data. This extended notation works for lots of commands, like `mean`, `sd`, `median`, `IQR`, `quantile`, etc. as long as the `mosaic` package is loaded.

### Your turn

Choose one numerical variable and one categorical variable from the `birthwt` data set. Find the five-number summary of your numerical variable grouped by your categorical variable.

\answerbegin
```{r}
# Add code here to compute the five-number summary of your
# numerical variable grouped by your categorical variable.
```
\answerend


## Conclusion

Summary statistics are simple numbers that describe and summarize data sets. Measures of center tell us where the "middle" of our data lies, and measures of spread tell us how spread out our data is. These measures should always be reported in pairs, for example the mean/standard deviation, or the median/IQR. Sometimes it can be useful to report summary statistics with the data separated by a categorical grouping variable.
